---
title: "Workshop_6_Assessment"
author: "Aiden Bennett"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
	echo = FALSE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```

### --- Introduction ---
This document contains the assessment items for Workshop 2 of Module 2. The Estuary Fish Survey uses four different data sets that need to be combined into one and provide a summary table and plot with abundance of the different species across salinity all at a publication standard.

### -- Phase 1: Data Cleaning --

Pre-Cleaning Actions
```{r, message=FALSE, warning=FALSE}
# Load packages
library(tidyverse)
library(stringr)
library(lubridate)
library(dplyr)
library(purrr)

# Loading the data
estuary_catch_log <-"C:/Documents/Masters Docs General/MB5370 - Techniques in Marine Science 1/Github/R4MarineScience/data/estuary_catch_log.xlsx"
sheets <- excel_sheets(estuary_catch_log)
catch <- map_dfr(
  sheets,
  ~ read_excel(estuary_catch_log, sheet = .x) %>%
    mutate(site = .x)
) # Combines the sheets of the estuary catch log into one data frame

# Renaming other data sets for easier use
species <- species_dictionary
sonde <- estuary_sonde_data
metadata <- estuary_metadata
```

First changes to the data formats
```{r, message=FALSE, warning=FALSE}
# Modify fish and site names in catch log to help match metadata and species dictionary
catch <- catch %>%
  mutate(
    site = site %>%
      str_trim() %>%
      str_to_lower() %>%
      str_replace_all("[[:space:]]+", "_"),
    species = species %>%
      str_trim() %>%
      str_to_lower() %>%
      str_replace_all("[[:space:]]+", "_")
  )

# Make sure meta data names match
metadata <- metadata %>%
  mutate(
    site_name = site_name %>%
      str_trim() %>%
      str_to_lower() %>%
      str_replace_all("[[:space:]]+", "_")
  )

# Fix date formats in sonde data, making it day, month, year, hour, minute format
sonde <- sonde %>%
  mutate(
    timestamp = dmy_hm(timestamp)
  )

# Changing incorrect values, -999, to NA in turbidity sensor 
sonde <- sonde %>%
  mutate(
    turbidity = na_if(turbidity, -999)
  )
```

### -- Phase 2: Relational Architecture --

Summarize water sensor data for temperature and salinity to mean daily values
```{r, message=FALSE, warning=FALSE}
sonde_daily <- sonde %>%
  mutate(day = floor_date(timestamp, unit = "day")) %>%
  group_by(site, day) %>%
  summarise(
    mean_temperature = mean(temperature, na.rm = TRUE),
    mean_salinity = mean(salinity, na.rm = TRUE),
    .groups = "drop"
  )
```

Merge fish species names from dictionary to catch data
```{r, message=FALSE, warning=FALSE}
catch <- catch %>%
   left_join(species, by = c("species" = "common_name")) %>%
  mutate(species = scientific_name) %>%
  select(-scientific_name)
```

Combine the data into one master frame
```{r, message=FALSE, warning=FALSE}
survey_summary <- catch %>%
  left_join(metadata, by = c("site" = "site_name")) %>%
  left_join(sonde_daily, by = c("site", "date" = "day"))
```


### -- Phase 3: Zero-Catch Framework --

Data was only recorded when a fish was caught, need to replace NA values with 0 and make a new framework. Adds rows where no fish were counted
```{r, message=FALSE, warning=FALSE}
zero_frame <- survey_summary %>%
  complete(site, species, date) %>%
  mutate(count = coalesce(count, 0))
```


### -- Phase 4: Statistical Extraction --

Make a clean summary table from the master sheet with mean and standard error of fish catch and salinity for each species in each zone
```{r, message=FALSE, warning=FALSE}
mean_values <- survey_summary





```